---
title: "Acute Lymphobastic Leukemia (ALL) GWAS"
author: "Abbas Rizvi"
date: "11/11/2018"
output: pdf_document  
---

```{r ch5_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

# Results

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/R_ALLonly_DRM.jpeg}
    \caption[Manhattan Plot of Recipient ALL DRM GWAS.]{Manhattan Plot of Recipient ALL DRM GWAS. Each dot is a SNP.}
    \label{fig:r_drm_1y}  
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/R_ALLonly_TRM.jpeg}
    \caption[Recipient TRM 1 Year Manhattan Plot]{Recipient TRM 1 Year Manhattan Plot}
    \label{fig:r_trm_1y}  
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/r_of_1y_gwas_hit.png}
    \caption[Recipient OF Region Chromosome 6.]{Recipient OF Region Chromosome 6. Sentinal SNP is rs6915410. Previous GWAS hit rs7744392 for trait contracts in type 2 diabetes. In perfect LD with rs13362808 which is cis eQTL in blood tissue.}
    \label{fig:r_of_1y}  
\end{figure}


\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/R_ALLonly_OS.jpeg}
    \caption[Recipient OS 1 Year Manhattan Plot]{Recipient OS 1 Year Manhattan Plot}
    \label{fig:r_os_1y}  
\end{figure}



```{r gwas_hits_1y, echo=FALSE}
gwas_hits <- readRDS("~/Google Drive/OSU_PHD/DBMT_ALLonly/hits/20190228_gwas_hits.rds")
gwas_hits %>%
    filter(str_detect(analysis, "1y")) %>%
    separate(analysis, c("genome", "disease", "outcome", "interval"), sep="_") %>%
    select(-interval, -disease, -query_snp_rsid) %>%
    mutate(genome=ifelse(genome=="D", "Donor", "Recipent")) %>%
    unite(analysis, c("genome", "outcome"), sep="\n") %>%
    rename(RSID=rsid,
           PMID=pmid,
           `GWAS Trait`=gwas_trait,
           `GWAS P-value`=gwas_pvalue,
           Analysis=analysis) %>%
    select(Analysis, everything()) %>%
    kable(caption="\\label{tab:gwas_hits_1y} Previously Published GWAS Traits from GWAS Catalog Reproduced in DISCOVeRY-BMT 1 Year ALL Analysis.", digits=60) %>%
    kableExtra::kable_styling(latex_options="striped") %>%
    kableExtra::column_spec(1, width="4em") %>%
    kableExtra::column_spec(2, width="5em") %>%
    kableExtra::column_spec(3, width="4em") %>%
    kableExtra::column_spec(4, width="8em") %>%
    kableExtra::column_spec(5, width="4em") %>%
    footnote(general_title="",
             general="SNPs shown here are either the query SNP or in LD (r2 > 0.8) with the query SNP.",
             footnote_as_chunk = TRUE,
             threeparttable = TRUE)


```

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/D_ALLonly_OS.jpeg}
    \caption[Manhattan Plot of Donor ALL OS GWAS.]{Manhattan Plot of Donor ALL OS GWAS. Each dot is a SNP. Green box is ...}
    \label{fig:d_os_1y}  
\end{figure}


\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/D_ALLonly_TRM.jpeg}
    \caption[Manhattan Plot of Donor ALL TRM GWAS.]{Manhattan Plot of Donor ALL TRM GWAS. Each dot is a SNP. This is here to talk about GVHD.}
    \label{fig:d_trm_1y}  
\end{figure}


For donor OS 1 year GWAS. rs12533724 imputed ($P = 1.51\text{e}^{-06}$; HR: 1.78, 95% CI 1.41-2.26).


rs62465751 (2b), rs7785755 (2b), rs10235334 (2c), rs62465735 (2b), rs28590087 (3a). This SNP is in LD with ..




\begin{figure}
    \centering
    \includegraphics[width=\linewidth, height=5in]{~/Desktop/figures/chapter5/rs9783053_1_D_ALLonly_GVHD.png}
    \caption[Donor GVHD Region Chromosome 1.]{Donor GVHD Region Chromosome 1. Sentinal SNP is rs9783053. RegulomeDB 1A. Typed SNP}
    \label{fig:d_gvhd_1y}  
\end{figure}




# Discussion