---
title: Genetic Associations in Acute Leukemia Patients after Matched Unrelated Donor
  Allogeneic Hematopoetic Stem Cell Transplantation
author: "Abbas A Rizvi, B.S., M.S., M.Sc."
documentclass: DissertateUSU
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
geometry:
- top=1in
- bottom=1in
- right=1in
- left=1.5in
- includefoot
params:
  committee1: Lara E Sucheston-Campbell, MS, PhD, Adviser
  committee2: Shili Lin, PhD
  committee3: Guy Brock, PhD
  committee4: Moray Campbell, PhD
  degree: Pharmaceutical Sciences
  gradschool: The Ohio State University
  year: '2019'
bibliography: dissertation_references.bib
---

```{r global_options, include=FALSE}
## This saves you from having to do this for every chunk
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
## Do NOT Remove
dissertateUSU::preamble_for_title()
```

<!-- Title page printed automatically -->
\pagenumbering{roman}
\pagestyle{empty}
\copyrightpage

<!-- Abstract -->
\newpage
\pagestyle{fancy}
\fancyhead[L]{Abstract}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ABSTRACT}
\addcontentsline{toc}{section}{Abstract}

```{r child = "abstract.Rmd"}
```


<!-- Dedication -->
\newpage
\fancyhead[L]{Dedication}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{DEDICATION}
\addcontentsline{toc}{section}{Dedication}

Dedicate it.

<!-- Acknowledgements -->
\newpage
\fancyhead[L]{Acknowledgments}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ACKNOWLEDGEMENTS}
\addcontentsline{toc}{section}{Acknowledgments}

```{r child = "acknowledgements.Rmd"}
```


<!-- Vita -->
\include{vita}


<!-- Front Matter -->
\newpage
\fancyhead[L]{Table of Contents}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\tableofcontents

\newpage
\fancyhead[L]{List of Tables}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\listoftables

\newpage
\fancyhead[L]{List of Figures}
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\listoffigures


\newpage
\pagenumbering{arabic}

<!-- Introduction -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter{Introduction}

\doublespacing

```{r child = "chapter1.Rmd"}
```

\FloatBarrier

<!-- Chapter 2 -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Survival Models}

```{r child = "chapter2.Rmd"}
```


\FloatBarrier

<!-- Chapter 3 -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Benchmarking survival model software}

```{r child = "chapter3.Rmd"}
```

\FloatBarrier
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Genome wide association study of Acute Lymphobastic Leukemia (ALL) Donors and Recipients after Transplant}

```{r child = "chapter4.Rmd"}
```

\FloatBarrier
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Conclusion}

```{r child = "chapter5.Rmd"}
```

\FloatBarrier

\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter*{REFERENCES}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.4in}
\setlength{\parskip}{6pt}
\noindent
