---
title: Genetic Associations in Acute Leukemia Patients after Matched Unrelated Donor
  Allogeneic Hematopoetic Stem Cell Transplantation
author: "Abbas A Rizvi"
documentclass: DissertateOSU
output:
  pdf_document:  
    includes:
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
header-includes:
- \usepackage{booktabs}
- \usepackage{placeins}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{amsrefs}
- \usepackage{filecontents}
- \usepackage{pdflscape}
- \usepackage{afterpage}
link-citations: yes
geometry:
- top=1in
- bottom=1in
- right=1in
- left=1.5in
- includefoot
params:
  committee1: Lara E Sucheston-Campbell, MS, PhD, Adviser
  committee2: Guy Brock, PhD
  committee3: Moray Campbell, PhD
  committee4: Shili Lin, PhD
  degree: Pharmaceutical Sciences
  gradschool: The Ohio State University
  year: '2019'
bibliography: dissertation_references.bib
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
## This saves you from having to do this for every chunk
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE,
                      cache=FALSE,
                      tidy=TRUE,
                      tidy.opts=list(width.cutoff=45))

## Do NOT Remove
library(tidyverse)
library(formatR)
library(kableExtra)

dissertateUSU::preamble_for_title()
```

<!-- Title page printed automatically -->
\pagenumbering{roman}
\pagestyle{empty}
\copyrightpage

<!-- Abstract -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\chapter*{ABSTRACT}
\addcontentsline{toc}{section}{Abstract}

```{r child = "abstract.Rmd"}
```


<!-- Dedication -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\chapter*{DEDICATION}
\addcontentsline{toc}{section}{Dedication}

Dedicated to Ezgi, my parents, my siblings and their kids, and my sweet little pooch Bernie. 

<!-- Acknowledgements -->
\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\chapter*{ACKNOWLEDGEMENTS}
\addcontentsline{toc}{section}{Acknowledgments}

\doublespacing

```{r child = "acknowledgements.Rmd"}
```


<!-- Vita -->
\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\chapter*{VITA}
\addcontentsline{toc}{section}{Vita}
\singlespacing

```{r child = "vita.Rmd"}
```


<!-- Front Matter -->
\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\tableofcontents

\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\listoftables

\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{}
\listoffigures

\newpage
\pagenumbering{arabic}

<!-- Introduction -->
\newpage
\pagestyle{plain} 
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter{Introduction}
\doublespacing

```{r child = "chapter1.Rmd"}
```

\FloatBarrier

<!-- Chapter 2 -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter{Replication and Validation of Previous HSCT Literature}
\doublespacing

```{r child = "chapter2.Rmd"}
```

\FloatBarrier

<!-- Chapter 3 -->
\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter{gwasurvivr: an R Bioconductor package for genome wide survival analysis}
\doublespacing

```{r child = "chapter3.Rmd"}
```

\FloatBarrier

<!-- Chapter 4 -->

\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter{Application and Pipeline}
\doublespacing

```{r child="chapter4.Rmd", eval=TRUE}
```


\FloatBarrier

<!-- Chapter 5 -->

\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Acute Lymphobastic Leukemia (ALL) GWAS}

\doublespacing

```{r child = "chapter5.Rmd", eval=TRUE}
```


\FloatBarrier

<!-- Chapter 6 -->

\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}

\chapter{Conclusion and Future Work}

\doublespacing

```{r child = "chapter6.Rmd", eval=TRUE}
```


\FloatBarrier

\newpage
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}


\chapter*{REFERENCES}
\singlespacing
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.4in}
\setlength{\parskip}{12pt}
\noindent

<div id="refs"></div>

\FloatBarrier

\newpage
\setlength{\parindent}{0.5in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{12pt}
\pagestyle{plain}
\fancyhead[L]{}
\fancyhead[R]{}
\fancyfoot[C]{\thepage}
\chapter*{APPENDICES}
\addcontentsline{toc}{chapter}{APPENDICES}

\doublespacing 

```{r child="ch2_appendix.Rmd", eval=TRUE}
```
